---
title: "Survival Analysis Project, Math 150"
author: "Siddhartha Aggarwal"
date: "4/8/2019"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, fig.height=3, fig.width=5,
fig.align = "center")
library(tidyverse)
library(broom)
library(dplyr)
library(ggplot2)
library(oilabs)
library(readr)

AIDSdata <- read_csv("~/AIDSdata.csv")



```


After uploading the data, the first visualization I wanted to create was to see how many individuals are part of each treatment group out of the 4 we have outlined. This would be interesting to get a sense of the variety of treatments in the study. The graph below shows that only Treatments 1 and 2 are administered in the study. This is odd, since not a single participant got treatments 3 or 4.

```{r}
tbl <- with(AIDSdata, table(txgrp))

barplot(tbl)


```



The next visualization I made was to plot the CD4 stratum at screening against the event indicator for AIDS defining diagnosis or death (censor variable). This was to get a sense of whether CD4 stratum levels in any way had a strong correlation with the censor variable.


```{r}

ggplot(AIDSdata, aes(x=jitter(strat2), y=jitter(censor))) + geom_point()

```


I made a few other visualizations but didn't really feel like they were useful enought to include. 


* * *














